Treatment heterogeneity in comparative effectiveness of teriparatide vs bisphosphonates: multi-database cohort study

### 2021-02-28 submission/manuscript.rmd
introduction citations

---
title: "Treatment heterogeneity in comparative effectiveness of teriparatide vs bisphosphonates: multi-database cohort study"
output:
    bookdown::pdf_document2: default
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
editor_options: 
  chunk_output_type: console
---

@@

\newpage
\linenumbers
# Introduction
Osteoporosis is a common condition characterised by decreased bone density and associated with increased risk for fragility fractures, which affects almost 30% of women aged $\geq50$ years (#Wright et al. 2017). It was estimated that in 2005, over 2 million incident fractures occurred in the United States, with hip fractures accounting for more than 70% of the costs. By 2025, annual fractures and costs could rise by almost 50% (#Cole et al. 2008).

Fracture prevention is thus the key focus of anti-osteoporotic therapy, reducing the disease burden both for the affected patient but also on the population level. Several pharmacological agents are available for primary and secondary prevention. The choice of anti-osteoporotic agent largely depends on history of fragility fracture and anticipated fracture risk. Oral bisphosphonates (BP) are first line treatments for postmenopausal patients with increased fracture risk in clinical guidelines (#AACE/ACE 2020) as for their favourable cost-effectiveness and safety profile. Teriparatide, a parathyroid hormone analogue administered as a daily injection, was approved by the FDA in 2002 as the first anabolic agent for treatment of severe postmenopausal osteoporosis. While clinical trials among patients with previous vertebral fractures showed a substantial risk reduction for new vertebral fracture comparing teriparatide to placebo (#Neer et al 2001) and risedronate (#Kendler et al. 2018), its efficacy on low-incident major osteoporotic fractures, especially hip fractures, is less well established: Previous randomised trials assessing hip fracture comprised only few events, thus not providing sufficient power to show differences between treatment groups when studying comparative effectiveness. A recent meta-analysis by Diez-Perez and colleagues assessed the effect of teriparatide on hip fracture, indicating a significant 80% risk reduction compared to placebo and a non-significant 46% risk reduction when compared to active controls.

With teriparatide-containing biosimilars being launched in recent years, treatment costs dropped significantly. Subsequently, the discussion of cost-effectiveness in patients with less severe forms of osteoporosis is restarting. Therefore, evaluation of the effectiveness in hip fracture prevention is required. In particular, observational studies assessing comparative effectiveness in a real-world setting are required, as effectiveness may differ from clinical trial as for differences in patient’s persistence and adherence to their anti-osteoporotic treatment. While hip fracture is a comparatively rare outcome, it is unambiguously defined and reliably recorded in routinely collected data, allowing for a direct comparison to results from clinical trial meta-analyses. While results from non-controlled, observational studies suggest that teriparatide may reduce the risk of hip fractures (#Burge et al. 2017, #Silverman et al. 2019), this is the first observational study assessing teriparatide in hip fracture risk reduction compared to oral bisphosphonate users in postmenopausal women using multiple large real-world databases.

@@

>>>>

---
title: "Treatment heterogeneity in comparative effectiveness of teriparatide vs bisphosphonates: multi-database cohort study"
output:
    bookdown::pdf_document2: default
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jamia.csl
---

@@

\newpage
\linenumbers
# Introduction
Osteoporosis is a common condition characterised by decreased bone density and associated with increased risk for fragility fractures, which affects almost 30% of women aged $\geq50$ years [@wright2017impact]. It was estimated that in 2005, over 2 million incident fractures occurred in the United States, with hip fractures accounting for more than 70% of the costs. By 2025, annual fractures and costs could rise by almost 50% [@Cole2008].

Fracture prevention is thus the key focus of anti-osteoporotic therapy, reducing the disease burden both for the affected patient but also on the population level. Several pharmacological agents are available for primary and secondary prevention. The choice of anti-osteoporotic agent largely depends on history of fragility fracture and anticipated fracture risk. Oral bisphosphonates (BP) are first line treatments for postmenopausal patients with increased fracture risk in clinical guidelines (#AACE/ACE 2020) as for their favourable cost-effectiveness and safety profile. Teriparatide, a parathyroid hormone analogue administered as a daily injection, was approved by the FDA in 2002 as the first anabolic agent for treatment of severe postmenopausal osteoporosis. While clinical trials among patients with previous vertebral fractures showed a substantial risk reduction for new vertebral fracture comparing teriparatide to placebo [@Neer2001] and risedronate [@Kendler2018], its efficacy on low-incident major osteoporotic fractures, especially hip fractures, is less well established: Previous randomised trials assessing hip fracture comprised only few events, thus not providing sufficient power to show differences between treatment groups when studying comparative effectiveness. A recent meta-analysis by Diez-Perez and colleagues assessed the effect of teriparatide on hip fracture, indicating a significant 80% risk reduction compared to placebo and a non-significant 46% risk reduction when compared to active controls [@DezPrez2019].

With teriparatide-containing biosimilars being launched in recent years, treatment costs dropped significantly. Subsequently, the discussion of cost-effectiveness in patients with less severe forms of osteoporosis is restarting. Therefore, evaluation of the effectiveness in hip fracture prevention is required. In particular, observational studies assessing comparative effectiveness in a real-world setting are required, as effectiveness may differ from clinical trial as for differences in patient’s persistence and adherence to their anti-osteoporotic treatment. While hip fracture is a comparatively rare outcome, it is unambiguously defined and reliably recorded in routinely collected data, allowing for a direct comparison to results from clinical trial meta-analyses. While results from non-controlled, observational studies suggest that teriparatide may reduce the risk of hip fractures [@Burge2016; @Silverman2018], this is the first observational study assessing teriparatide in hip fracture risk reduction compared to oral bisphosphonate users in postmenopausal women using multiple large real-world databases.

@@

\newpage
# Discussion

\newpage
# References
\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent

<div id="refs"></div>

\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\noindent
